---
title: "World Happiness Case Study"
author: "Jace Wilderman"
date: "2025-07-06"
output: html_document
---

## Background information
This case study uses information from the World Happiness Report to examine the relationship between the happiness score of a country or region and several different factors.
During this case study, we primarily use the 2019 World Happiness Report for our data. 
However, at the end we will look at the score changes between 2018 and 2019 and see how each factor has changed as well.


## Install and load 'tidyverse' packages (optional)

If `tidyverse` has not already been installed and loaded, run the code segment below to ensure that there will not be issues when calling functions included with the `tidyverse` package.

```{r installing and loading tidyverse, echo=FALSE, message=FALSE}
library(tidyverse)  
```

## Importing Data

Run the code below to load the 2019 World Happiness Report `2019.csv` to a data frame.
The rename function is also used to change the names of the columns.

```{r load 2019 data and rename columns, echo=FALSE, message=FALSE}
happiness_2019 <- read_csv("2019.csv")

happiness_2019 <- rename(happiness_2019, 
                         Happiness_Rank = "Overall rank",
                         Country = "Country or region",
                         Score = Score,
                         Economy = "GDP per capita",
                         Social_support = "Social support",
                         Life_expectancy = "Healthy life expectancy",
                         Freedom_of_choice = "Freedom to make life choices",
                         Generosity = Generosity,
                         Perceptions_of_corruption = "Perceptions of corruption")
```

## Plotting correlations

With this first chart, we plot how Economy correlates to Score to see if there are any trends that can be defined.
We include a trendline to highlight the trend we can see the plots following.

```{r Plotting how Economy correlates to Score, echo=FALSE}
ggplot(data = happiness_2019, mapping = aes(x = Score, y = Economy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

With this second chart, we view the correlation between Social_support and Score.

```{r Plotting how Social_support correlates to Score, echo=FALSE}
ggplot(data = happiness_2019, mapping = aes(x = Score, y = Social_support)) +
  geom_point() +
  geom_smooth(method = "lm")

```

This third chart shows the correlation that Life_expectancy has on Score.

```{r Plotting how Life_expectancy correlates to Score, echo=FALSE}
ggplot(data = happiness_2019, mapping = aes(x = Score, y = Life_expectancy)) +
  geom_point() +
  geom_smooth(method = "lm")
```

With each of these charts, we can see that each of these factors play some part in the happiness score of a country.

## Importing data for 2018 World Happiness Report

To gain a better view on how the changes in these factors affect happiness scores.

```{r load 2018 data and viewing dataframe structures, echo=FALSE, message=FALSE}
happiness_2018 <- read_csv("2018.csv")

str(happiness_2019)
str(happiness_2018)
```

There are some differences in data structure and column names, so we will address those before we use the data frame. 
After running this code segment, we will get a warning, but this is expected and will not cause issues.

```{r rename columns and change data structure, echo=FALSE, message=FALSE}
happiness_2018 <- rename(happiness_2018, 
                         Happiness_Rank = "Overall rank",
                         Country = "Country or region",
                         Score = Score,
                         Economy = "GDP per capita",
                         Social_support = "Social support",
                         Life_expectancy = "Healthy life expectancy",
                         Freedom_of_choice = "Freedom to make life choices",
                         Generosity = Generosity,
                         Perceptions_of_corruption = "Perceptions of corruption")

happiness_2018$Perceptions_of_corruption <- as.numeric(happiness_2018$Perceptions_of_corruption)
```

## Combining data and calculating score changes

Now we will combine the data from both `happiness_2018` and `happiness_2018` by Country.

```{r merging data frames, echo=FALSE, message=FALSE}
happiness_merged <- inner_join(happiness_2019, happiness_2018, by = "Country")
```

With this combined data frame will calcuate the score differences between the year and save them to a new data frame.

```{r calculate score differences, echo=FALSE, message=FALSE}
happiness_change <- transmute(happiness_merged, Country = Country,
                              Score_change = Score.x - Score.y, 
                              Economy_change = Economy.x - Economy.y, 
                              Social_support_change = Social_support.x - Social_support.y,
                              Life_expectancy_change = Life_expectancy.x - Life_expectancy.y,
                              Freedom_of_choice_change = Freedom_of_choice.x - Freedom_of_choice.y,
                              Generosity_change = Generosity.x - Generosity.y,
                              Perceptions_of_corruption_change = Perceptions_of_corruption.x - Perceptions_of_corruption.y)
```

## Conclusions based on score changes

We will now plot the new data changes to see if we can identify correlations.

```{r}
ggplot(data = happiness_change, mapping = aes(x = Score_change, y = Economy_change)) +
  geom_point() +
  geom_smooth(method = "lm")
```

```{r}
ggplot(data = happiness_change, mapping = aes(x = Score_change, y = Social_support_change)) +
  geom_point() +
  geom_smooth(method = "lm")
```

When looking at these charts we can notice that there is not as strong of a correlation between each factor and the change in score.

## Conclusions

1. Each factor does influence the score, but if we want to see a greater increase in happiness score, we must work to improve each factor not just focusing on one.
2. Efforts can be more efficient if directed towards areas that the country has scored lower in.
3. Just because a country has a lower happiness score, it does not mean that all of their scores are lower. It might mean that a few of these areas might need to be improved on.
